Optimal planning of factorial experiments in case of model contamination

### 2021-05-27 MSE_paper.tex
Version 27/05/2021

\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}

@@

\begin{document}
\tableofcontents

@@

\textit{
	Model-independent approach to building designs in case of model uncertainty
	\begin{itemize}
		\item[-] Planning a controlled intervention -- reliance on a model (1) +
		\item[-] Active assumption of a specific model misspecification (2) =
		\item[-] Combining conflicting aims: model-based inference based on the model-independent variance estimate and distrust in the model sufficiency
		\item[-] Briefly: extension on more complex experimental structures
	\end{itemize}
}

Designed experiments are commonly conducted in order to gain understanding of the impact that different parameters of interest make on the outcome; the quantitative measure of this impact allows comparisons as well as making interpretable conclusions regarding the scale and the pattern of the relationships between process parameters and measured output.

@@

Whichever the chosen, `primary',  model is, planning a controlled intervention does rely on the approximating model assumptions in two main -- and quite contradicting -- directions(?). Firstly, it is desirable to ensure the goodness of what the model is essentially fitted for, that is, for example, the precision of model parameters and/or the prediction accuracy of the untested treatment combinations. The corresponding design aims are usually reflected in utilising well-known optimality criteria (D-, A-, ?prediction) and rely on the variance estimate, the way it is obtained. Secondly, treating the chosen model as the absolute truth, especially at the designing stage is at least too optimistic and, as in many cases fitting a more suitable, but complicated model might not be feasible due to various restrictions and limitations, could be dangerous for the results credibility. So having a particular model also means that at the stage of planning it is highly desirable to include some [sensitivity towards and robustness against]  control over the model lack of fit and its effect on the inference.

We deal with such duality [of model-dependence and accounting for its misspecification] by developing compound optimality criteria, each constructed as a weighted combination of individual criterion functions, with the two main features:
\begin{enumerate}
	\item Each corresponds to a specific desirable property:  either accounting for an aim coming from trusting the model or mitigating the effects from its potential misfit. The relative importance of the components are reflected by the assigned weights -- and we shall examine the performance of the resulting designs in terms of the individual criteria, and explore the role of weight allocation.
	\item The concept of model-independent internal variance estimation, `pure error' (\cite{GilmourTrinca2012})) underlies each of the individual criteria -- the most appropriate [sensible] strategy in the case of possible model insufficiency. 
 \end{enumerate}

The general spirit of this work aims at aligning with the concept of a good design, which, as summarised by \cite{Box1987empirical},  should ``make it possible to detect lack of fit'' and ``provide an internal estimate of error from replication'', among other properties.

We will focus on factorial experiments with a relatively small number of runs and the fitted model being a polynomial regression. Section \ref{sec::background} provides the background on the modelling, error estimation and fundamental individual criteria. Controlling the lack-of-fit and the bias arising from the model misspecification are introduced in Section \ref{sec::criteria}, where they are combined with the primary model-driven ones in compound optimality criteria. Their adaptation to experimental frameworks with restricted randomisation is described in Section \ref{sec::ms_experiments}, as well as the design construction procedure for the multistratum experiments. Finally, a series of examples are presented in Section \ref{sec::examples} -- examining dynamics across various optimal designs,  shape of the constructed criteria and other properties, followed up by a Discussion in Section \ref{sec::discussion} with main conclusions and recommendations.

The $\mbox{R}$ code used to search for the optimal designs are located at <...> [?github for now].

@@

\textit{	
	\begin{itemize}
		\item[-] Modelling and relevant error estimation 
		\item[-]  Fundamental individual criteria
		\item[-] Combining objectives: lit.review
	\end{itemize}
}

@@

\begin{equation}
\label{eq::back_model}
\bm{Y}=\bm{X\beta}+\bm{\varepsilon}.
\end{equation}

@@

Any inference following the model fitting relies on the error variance estimate  $\hat{\sigma}^2$. The most common one is the mean square error: $\hat{\sigma}^2_{mse}=\mbox{Residual SS}/(n-p)$ (e.g. \cite{Draper1998}), which is model-dependent with the residual degrees of freedom $\nu = n-p$ containing the number of model parameters.

@@

\subsection*{Optimality criteria}

Model-dependency at the stage of experimental planning is reflected in searching for a design that optimises a criterion function, that is a function of a design that translates a specific inference-driven objective. For example, among the most well-known, ``alphabetic'' optimality criteria, such as $D$-, $C$-, $L$-optimality and a series of others (\cite{Atkinson2007}) target the precision of parameters estimates in model (\ref{eq::back_model}); while others, like $G$- and $V$-optimality, deal with the prediction variance. Traditionally these criteria are formulated based on the mean square error estimate; citet{GilmourTrinca2012} derived the alternative pure-error based criteria, which would guarantee the presence of replicates in the resulting designs.

Combining multiple desirable objectives in the design can be fulfilled through constructing a compound criterion. This concept is based on the notion of design efficiency, which can be defined for any design matrix $X$ and any criterion $F(X)$ as the ratio with respect to the best (e.g. maximum, without the loss of generality) value achieved at the optimal design. For example, the $D$-efficiency of design $\bm{X}$ is
\begin{equation*}
%\label{eq::D_eff}
\mbox{Eff}_{D}(X)=\frac{\vert \bm{X}'\bm{X}\vert^{1/p}}{\vert \bm{X}'_{*}\bm{X}_{*}\vert^{1/p}},
\end{equation*}   
where $\bm{X}_{*}$ is the $D$-optimum design. In this definition the power $1/p$ brings the efficiency to the scale of variances of model coefficients $\bm{\beta}_{i}, i=1\ldots p.$ The efficiency value may vary from $0$ to $1$ and is equal to $1$ if and only if the design is optimal according to the criterion of interest.

@@

Some of the alternative approaches of combining several objectives would include generating the Pareto optimal set of designs -- the approach thoroughly described by \cite{Lu2011optimization}.  \cite{Stallings2015general}developed methodology for generalising eigenvalue-based criteria (e.g. $A$- and $E$-optimality) in a way that allows differing interest (expressed through the weights) among any set of estimable functions of the fitted model parameters. The introduced strategy reflects the aims of experimentation that are not traditionally accounted for but definitely are of interest. [Add more references to compound optimality?]

\section{Individual criteria construction}
\label{sec::criteria}
\textit{	
	\begin{itemize}
		\item[-] Model misspecification setup. What objectives to be reflected in criteria: PE
		\item[-] Lack-of-fit criterion (and then the traditional LoF)
		\item[-] MSE criterion
		\item[-] Combining with DP/LP -- the general compound criterion
	\end{itemize}
}
\input{compound_criterion}

\section{Restricted randomisation}
\label{sec::restricted_randomisation}
\input{restricted_randomisation}

\section{Examples}
\label{sec::examples}
\input{compound_examples}

%% Criteria for blocked experiments as a special case of MS

@@

\section{Discussion}
\label{sec::discussion}

@@

%%% Bibliography
\cleardoublepage
\phantomsection
\addcontentsline{toc}{chapter}{Bibliography} 
%\bibliographystyle{apalike}
\bibliographystyle{rss}
\bibliography{thesis_bib}

>>>>

\newcommand{\highlight}{\textcolor{red}}
\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}

@@

\begin{document}

\title{Optimal planning of factorial experiments in case of model contamination}
\author{\large{Olga Egorova, Steven G. Gilmour}}
\date{[Draft]}
\maketitle

\tableofcontents

@@

Designed experiments are commonly conducted in order to gain understanding of the impact that different parameters of interest make on the outcome; the quantitative measure of this impact allows comparisons as well as making interpretable conclusions regarding the scale and the pattern of the relationships between \highlight{experimental factors} and measured output.

@@

Whichever the chosen, `primary',  model is, planning a controlled intervention does rely on the approximating model assumptions in two main -- and quite contradicting -- directions. Firstly, it is desirable to ensure the goodness of what the model is essentially fitted for, that is, for example, the precision of model parameters and/or the prediction accuracy of the untested treatment combinations. The corresponding design aims are usually reflected in utilising well-known optimality criteria ($D$-, $A$-, $G$-,$V$-, etc.), which rely on the variance estimate, making the way it is obtained crucial. Secondly, treating the chosen model as the absolute truth, especially at the designing stage is at least too optimistic and, as in many cases fitting a more suitable, but complicated model might not be feasible due to various restrictions and limitations, could be dangerous for the results credibility. So having a particular model also means that at the stage of planning it is highly desirable to include some control over the model lack of fit and its effect on the inference.

We deal with such duality of model-dependence and accounting for its misspecification by developing compound optimality criteria, each constructed as a weighted combination of individual criterion functions, with the two main features:
\begin{enumerate}
	\item Each corresponds to a specific desirable property:  either accounting for an aim coming from trusting the model or mitigating the effects from its potential misfit. The relative importance of the components are reflected by the assigned weights -- and we shall examine the performance of resulting optimal designs in terms of the individual criteria, and explore the role of weight allocation.
	\item The concept of model-independent internal variance estimation, `pure error' (\cite{GilmourTrinca2012})) underlies each of the individual criteria -- the most appropriate and sensible strategy in the case of possible model insufficiency. 
 \end{enumerate}

The general spirit of this work aims at aligning with the concept of a good design, which, as summarised by \cite{Box1987empirical},  should ``make it possible to detect lack of fit'' and ``provide an internal estimate of error from replication'', among other properties. Classical designs, like central composite ones (CCDs, \cite{Myers2009}), have been very popular in practice due to their geometrical properties.

We will focus on factorial experiments with a relatively small number of runs and the fitted model being a polynomial regression. Section \ref{sec::background} provides the background on the modelling, error estimation and fundamental individual criteria. Controlling the lack-of-fit and the bias arising from the model misspecification are introduced in Section \ref{sec::criteria}, where they are combined with the primary model-driven ones in compound optimality criteria. Their adaptation to blocked experiments is described in Section \ref{sec::compound_blocked}, followed by the optimal design construction and variance analysis for experimental frameworks with restricted randomisation in Section \ref{sec::multistratum_experiments}.
A series of examples are  presented -- examining dynamics across various optimal designs,  shape of the constructed criteria and other properties, followed up by a Discussion in Section \ref{sec::discussion} with main conclusions and recommendations.

The $\mbox{R}$ code used to search for the optimal designs are located at [github/R package link].

@@

@@

\begin{equation}
\label{eq::primary_model}
\bm{Y}=\bm{X\beta}+\bm{\varepsilon}.
\end{equation}

@@

Any inference based on building confidence regions and hypothesis testing following the model fitting relies on the error variance estimate $\hat{\sigma}^2$. The most common one is the mean square error: $\hat{\sigma}^2_{mse}=\mbox{Residual SS}/\nu$ (e.g. \cite{Draper1998}), which is model-dependent with the residual degrees of freedom $\nu = n-p$ containing the number of model parameters.

@@

%\subsection*{Optimality criteria}
However, model-dependency at the stage of experimental planning is reflected in searching for a design that optimises a criterion, that is a function of a design that translates a specific inference-driven objective. For example, among the most well-known, ``alphabetic'' optimality criteria, such as $D$-, $C$-, $L$-optimality and a series of others (\cite{Atkinson2007}) target the precision of parameters estimates in model (\ref{eq::back_model}); while others, like $G$- and $V$-optimality, deal with the prediction variance; traditionally these criteria are formulated based on the mean square error estimate. \citet{GilmourTrinca2012} derived the alternative, model-independent pure-error based criteria, which would guarantee the presence of replicates in the resulting designs. \highlight{Primary criteria were formulated for interval-based inferential properties: minimising the volume of a $(1-\alpha_{DP})\times 100\%$ confidence region for the model parameters ($DP$-optimality) or the mean squared lengths of the $(1-\alpha_{LP})\times 100\%$ confidence intervals for linear functions of the parameters' estimates' variances ($LP$-optimality):
\begin{align}
\label{DP_LP}
DP: \mbox{  }(F_{p,d;1-\alpha_{DP}})^{p}\vert(\bm{X'}\bm{X})^{-1}\vert &\longrightarrow \mbox{ min}, \\ \notag
LP: \mbox{ }F_{1,d;1-\alpha_{LP}}\mbox{tr}\{\bm{W}(\bm{X}'\bm{X})^{-1}\} &\longrightarrow \mbox{ min},
\end{align}
where $F_{df_1,df_2;1-\alpha}$ is the ``upper $\alpha$-point'' of the F-distribution with $df_1$ and $df_2$ numerator and denominator degrees of freedom.
}

Combining multiple desirable objectives in the design can be fulfilled through constructing a compound criterion. This concept is based on the notion of design efficiency, which can be defined for any design matrix $X$ and any criterion $C(X)$ as the ratio with respect to the best (e.g. \highlight{minimum}, without the loss of generality) value achieved at the optimal design. \highlight{For example, the $DP$-efficiency of design $\bm{X}$ is
\begin{equation*}
%\label{eq::DP_eff}
\mbox{Eff}_{DP}(X)=\frac{\vert \bm{X}'_{*}\bm{X}_{*}\vert^{1/p}/F_{p,d_{*};1-\alpha_{DP}}}{\vert \bm{X}'\bm{X}\vert^{1/p}/F_{p,d;1-\alpha_{DP}}},
\end{equation*}   
where $\bm{X}_{*}$ is the $DP$-optimum design with $d_{*}$ pure error degrees of freedom.} In this definition the power $1/p$ brings the efficiency to the scale of variances of model coefficients $\bm{\beta}_{i}, i=1\ldots p.$ The efficiency value may vary from $0$ to $1$ and is equal to $1$ if and only if the design is optimal according to the criterion of interest.

@@

Some of the alternative approaches of combining several objectives would include generating the Pareto optimal set of designs -- the approach thoroughly described by \cite{Lu2011optimization}.  \cite{Stallings2015general}developed methodology for generalising eigenvalue-based criteria (e.g. $A$- and $E$-optimality) in a way that allows differing interest (expressed through the weights) among any set of estimable functions of the fitted model parameters. The introduced strategy reflects the aims of experimentation that are not traditionally accounted for but definitely are of interest.

\section{Criteria Accounting for Model Uncertainty}
\label{sec::criteria}
\input{compound_criterion}

\subsection{Example}
\label{sec::examples}
\input{compound_examples}

\section{Blocked Experiments}
\label{sec::compound_blocked}
\input{compound_blocked}

@@

\section{Multistratum Experiments}
\label{sec::multistratum_experiments}
\input{multistratum_experiments}

\subsection{Split-split-plot optimal design}
\label{subsec::ms_example}
\input{ssp_example}

\section{Discussion}
\label{sec::discussion}
\textit{
\begin{itemize}
	\item Conclusions
	\item Focus on future work(?) and/or alternatives
\end{itemize}
}
%The adapted MSE-based compound criteria, error estimation and design search procedure brought together allows for providing compromise across competing objectives in a complex experimental framework -- and tools for a decision making reflecting the priorities and aims of the experimentation.

@@

\clearpage

\appendix
\section{Appendices}
\subsection{Appendix I}
\label{appendix::example1}
\input{appendix_example}

\subsection{Appendix II}
\label{appendix::ms_df}
\input{appendix_df}

\subsection{Appendix III}
\label{appendix::ms_design}
\input{appendix_ms_design}

%%% Bibliography
\cleardoublepage
\phantomsection
%\addcontentsline{toc}{chapter}{Bibliography} 
%\bibliographystyle{apalike}
\bibliographystyle{rss}
\bibliography{thesis_bib}

### 2021-05-27 appendix_ms_design.tex
Version 27/05/2021

\textit{	
	\begin{itemize}
		\item[-] General framework of multistratum experiments. [add more on model uncertainty]
		\item[-] Error estimation: REML and pure error REML. Yates procedure
		\item[-] Adapting optimality criteria and design construction
	\end{itemize}
}

In a large number of industrial, engineering and laboratory-based experiments, either the nature of the study or certain restrictions result in the necessity of considering a multi-level structure of experimental units. For example, a chemical process consisting of applying treatments to the material batches of different sizes at each stage;  or one or several experimental factors' values can be changed only once per a certain amount of runs whereas values of other factors are varied between runs. Therefore, different factors are applied at different levels, and randomisation is performed at each level, thus the whole process of allocating treatments to experimental units should be amended accordingly \citep{MeadGilmour2012}.

In this work we go on to consider experimental framework comprising such a hierarchical structure of experimental units and treatments, which is referred to as multistratum experiment, and each stratum is defined as a level in the unit structure. Units are grouped into whole-plots, each of them divided into sub-plots, which contain a certain number of sub-sub-plots, and so on up to the smallest units --- runs of the experiment -- [Figure .with Hasse diagram?] . In the general case of two strata, we deal with what is called a ``split-plot'' experiment, in case of three strata a ``split-split-plot'' experiment. 
%\begin{figure}[h]
%	\begin{center}
%		\includegraphics[scale=0.85]{Hasse.jpg}      %width=\textwidth
%		\caption{Hasse diagram for factors}
%		\label{Fig::Hasse}
%	\end{center}
%\end{figure} 
Our aim is to adapt the MSE-based criteria derived before to the factorial experiments with units distributed in several strata. The most important aspects are (1) estimating the error components at all of the levels and (2) formulating the compliant criteria and provide the suitable implementation strategy for the optimal design search.

\subsection{Response-Surface Methodology for multistratum framework}
%% Model. Ratio of variances.
[OE: introduce the notion of "block"?]
Here we are considering experiments with $s$ strata in total, stratum $i$ being nested within units of the stratum $i-1$, and stratum $0$ will be seen as the whole experiment (following the notation set by \citet{Trinca2015improved}). The number of units in stratum $i$ within every unit in stratum $i-1$ is denoted by $n_i$, such that $m_{j}=\prod_{i=1}^{j}n_{i}$ is the number of units in stratum $j$ and, therefore,  $n=m_{s}=\prod_{i=1}^{s}n_{i}$ is the total number of runs. As randomisation occurs at each stratum, the model accounting for the hierarchical error structure and correlated observations can be written as
\begin{equation}
\label{eq::MS_model}
\bm{Y}=\bm{X}\bm{\beta}+\sum_{i=1}^{s}\bm{Z}_{i}\bm{\varepsilon}_{i},
\end{equation}
where $\bm{Y}$ represents the $n$-dimensional vector of responses, $\bm{X}$ stands for the $n \times p$ model matrix and $\bm{\beta}$ for the $p$-dimensional vector of corresponding model coefficients. Each row of the $n\times m_{i}$ matrix $\bm{Z}_{i}$ corresponds to a single experimental run and indicates the unit in stratum $i$ containing this run. $\bm{\varepsilon}_{i}$ is a vector of random effects occurring due to the randomisation at level $i$, and these effects are assumed to be independent and identically distributed around zero mean and variance $\sigma^{2}_{i}$. Generalised Least Squared  (GLS) estimators of $\bm{\beta}$ are calculated as
\begin{equation}
\label{eq::MS_GLS}
\bm{\hat{\beta}}=(\bm{X}'\bm{V}^{-1}\bm{X})^{-1}\bm{X}'\bm{V}^{-1}\bm{Y}.
\end{equation}

%The assumption of error normality, though quite common is required for some analysis methods, is not a necessary one at this stage.

%Inferential purposes of obtaining parameter estimates $\bm{\hat{\beta}}$ of good quality imply the necessity of evaluating the dispersion parameters $\sigma^2_{i}$ that are directly related to the estimates' variance and, therefore, responsible for measuring the uncertainty of the inference in a general sense.

\subsubsection*{Error Estimate}
An extensive amount of research has been conducted on the design of and analysis of data from split-plot and split-split-plot experiments, with one of the first works by \cite{Letsinger1996BiRandomization}, emphasising the necessity of adapting the design and analysis strategy with respect to the more complicated error structure.  In the most general case of non-orthogonal unit structures Residual Maximum Likelihood (REML) methodology has been acknowledged to be the most sensible approach to estimating the variance components by maximising the part of the likelihood function corresponding to the ``random'' part of the model; the details are provided, for example, in \cite{Searle2001generalized}.

%In traditional REML-based methodology the estimates of $\sigma^{2}_{i}$ are obtained from model (\ref{eq::MS_model}) and then GLS estimators of $\bm{\beta}$ are calculated as
%\begin{equation}
%\label{eq::MS_GLS}
%\bm{\hat{\beta}}=(\bm{X}'\bm{V}^{-1}\bm{X})^{-1}\bm{X}'\bm{V}^{-1}\bm{Y},
%\end{equation}
%where $\bm{V}$ is the unknown variance-covariance matrix of the response vector, to which the estimates are to be plugged in:
%\begin{equation*}
%\bm{V}=\sum_{i=1}^{s}\sigma^{2}_{i}\bm{Z}_{i}\bm{Z}'_{i},
%\end{equation*}
%so that $\mbox{Var}(\bm{\hat{\beta}})=(\bm{X}'\bm{V}^{-1}\bm{X})^{-1}.$

In some case, though,  REML tends to underestimate the variance components in the higher stratum, for example \cite{Goos2006practical} considered some split-plot experiments with true whole-plot variances being non-negligibly far from zero, for which REML, however, provided zero estimates. A possible alternative of following Bayesian strategy \citep{Gilmour2009analysis}, however, this requires a careful choice of the prior which is often not possible, especially in the case of more than two strata.

In the context of model uncertainty, it would be appropriate to consider instead the direct relationship between the response and the set of treatments, similar to (\ref{eq::treatment_model}):
\begin{equation}
\label{eq::treatmentMS}
\bm{Y}=\bm{X_t}\bm{\mu_t}+\sum_{i=1}^{s}\bm{Z}_{i}\bm{\varepsilon}_{i}.
\end{equation}

Each column of $\bm{X_t}$ corresponds to a treatment defined as a combination of the experimental factors;
%in case of factorial experiment the total number of possible treatments (or candidate design points) $T$ would be the product of the number of the factors' levels. 
$\bm{Z}_{i}$ is the indicator matrix of random effects at stratum $i$.

\cite{GilmourGoos2016PEREML} introduced the notion of ``Pure Error REML'', where random effects are estimated from the treatment model (\ref{eq::treatmentMS}), and then for the sake of making inference regarding parameters of the response model (\ref{eq::MS_model}), they are substituted in the GLS estimators of the fixed effects (\ref{eq::MS_GLS}). It is also argued that some appropriate corrections are to be adapted and applied to the obtained estimates, i.e.~the ones suggested by \cite{kenward1997small}.

A careful definition of ``pure error'' is necessary when considering the treatment and response surface models, especially in the presence of nested random block effects. \cite{Gilmour2000PErsm} discuss the pure error estimation issues in the context of blocked experiments.

[OE: The following is to be paraphrased -- made much shorter or removed overall?]
Pure error is expected to measure the variability between experimental units regardless of the treatments applied, and therefore, the assumption of treatment-unit additivity is essential in unblocked experiments.  In the presence of one or several blocking factors, the existence of block-treatment interaction would imply that pure error in the lower strata can only be estimated from the replicates within blocks, so that inter-block information is not taken into consideration. Hence, in the blocked cases the assumption of treatment-block additivity \citep{Draper1998} is desirable. \cite{Gilmour2000PErsm} argue for the preservation of treatment-unit additivity when either fixed or random block effects are in the model in order to preserve consistency with the unblocked experiments; that would also conveniently imply treatment-block additivity. Adoption of these assumptions would then lead to the following representation of the relationship between responses and treatments in multi-stratum experiments.

The outcome of applying treatment $i$ $(i=1\ldots T)$ to the experimental run located in the units $j_1$ of the first stratum, $j_2$ of the second stratum, $\ldots$ , and unit $j_s$ of the $s-$th stratum can be expressed as follows:
\begin{equation}
\label{eq::ms_tr}
y_{ij_{1}...j_{s}}=\mu+t_{i}+b_{j_1}+\ldots +b_{j_s},
\end{equation}
where $b_{j_s}$ is usually denoted as between-run variation.

Under the stronger assumption of the response surface model (\ref{eq::MS_model}), the treatment effects $t_{i}$ are represented as a set of polynomial model terms.

Based on the interrelation of the two models, and the desirability of being able to provide the possibility for testing for the lack-of-fit together with obtaining robust estimates of the variance components, the approach of using pure error is adopted, \cite{GilmourGoos2016Robust} explored several approaches to estimating the variance components. \\[5 pt]
[OE: the end of the part to be re-written/excluded]

The first approach is stratum-by-stratum data analysis, where each randomisation level is considered separately, and at each stratum $i$ units are treated as runs aggregated in $m_{i-1}$ blocks with fixed effects. The lower stratum variance is estimated from within blocks (using intra-block residual mean square), and the higher stratum (inter-block) variance from the difference between the intra- and inter-block residual mean squares scaled by the number of runs per block \citep{Hinkelmann2005Advanced}. This strategy requires only the treatment-unit additivity assumption and provides unbiased treatment effects' estimators. However, it would be desirable to make use of combining information not just from within-strata (intra-block) treatment replicates, but also from between-strata (inter-block) ones, especially in the common cases of relatively small experiments when the amount of runs available does not allow for obtaining higher stratum variance components from only considering blocks as whole units. Yates' procedure, first suggested by \cite{yates1939recovery} and described in detail by \cite{Hinkelmann2005Advanced}, provides more degrees of freedom for the variance estimate in the higher stratum -- and this is the approach we are adopting in this work.

In the case of two levels of randomisation, variance components are estimated in two variance decomposition steps: 
\begin{enumerate}
	\item Total SS = SS(Blocks) + SS(Treatments$\vert$Blocks) + SS(Residual)\\
	From fitting this full treatment-after-blocks model the residual mean square $S^2$ is taken as the estimate of the intra-block variance. SS(Residual) is then substituted in the following partition.
	\item Total SS = SS(Treatments) + SS(Blocks$\vert$Treatments) + SS(Residual)\\
	Therefore, the sums of squares for the blocks after the treatment effects have been accounted for is obtained and the corresponding mean square $$S^2_{b}=\mbox{SS(Blocks$\vert$Treatments)}/\nu_{b}$$ is then used to get an estimate of the inter-block variance:
	$\hat{\sigma}^2_{b}=\frac{\nu_{b}(S^2_{b}-S^2)}{d}$,
	where pure error degrees of freedom for inter-block variance is $d=n-\mbox{trace}[\bm{Z}'\bm{X_t}(\bm{X_t}'\bm{X_t})^{-1}\bm{X_t}'\bm{Z}]$, $\nu_{b}=\mbox{rank}([\bm{X_t} \bm{Z}])-\mbox{rank}(\bm{X_t})$, and $\bm{X_t}$ and $\bm{Z}$ are as in (\ref{eq::treatmentMS}).
\end{enumerate}

Allocation of the degrees of freedom in practice, following Yates' procedure, is provided in the Appendix.

[OE: to shorten]\\
\cite{GilmourGoos2016Robust} show that in addition to the treatment-unit additivity and randomisation the assumption of the experimental units being a random sample from an infinite population is necessary for the inter-block variance estimate to be unbiased. In this particular context, when the normality and independence of responses in (\ref{eq::MS_model}) is a standard assumption, although a potential presence of contamination effects in the fixed part of the model is also accounted for, Yates' approach seems to be the most appropriate technique. However, relying on the distribution assumption to the extent that makes REML the most sensible approach means that the fixed part of the fitted model is assumed to be absolutely true, that is the parameters of the population distribution are known, which does not comply with the model uncertainty framework.

Together with the specifications regarding estimation of the variance components, it is necessary to carefully establish the way the pure error degrees of freedom should be determined and how the available remaining number of treatment degrees of freedom would be distributed between the polynomial, lack-of-fit and inter-block information components.

\subsection{MSE-based compound criteria for multistratum experiments}
\label{sec::ch7_search}
In the presence of potential model disturbance which is expressed as additional polynomial terms, the full model for a multistratum experiment is then:
\begin{equation}
\label{eq::MS_model_full}
\bm{Y}=\bm{X}_p\bm{\beta}_p+\bm{X}_q\bm{\beta}_q+\sum_{i=1}^{s}\bm{Z}_{i}\bm{\varepsilon}_{i},
\end{equation}
where $\bm{X}_p$ is the primary terms matrix, and $bm{X}_q$ contains potential terms.

The design construction will follow the approach developed by \cite{Trinca2015improved}: an iterative, stratum-by-stratum algorithm, that treats the higher stratum units as fixed blocks; such an approach does not require any prior assumptions regarding the values of the variance components.

At each step a candidate set of treatments applied at the current stratum is to be generated, and the point exchange algorithm is applied. The model matrix to be used in the optimisation comprises model terms from all higher strata (with the factor values for individual runs obtained at previous steps), terms from the current stratum and interactions between these two groups of terms, if there are any to be considered. The allocation of the degrees of freedom shall be implemented according to the Yates procedure, and some illustrative examples will be given later on.

\subsection*{Criteria for blocked experiments}
\label{subsec::compound_blocked}
\input{compound_blocked}

%Before moving to a more detailed presentation of the design construction algorithm, recall the MSE-based criteria (\ref{eq::MSE_D}), (\ref{eq::MSE_L}), (\ref{eq::MSE_D_B}) and (\ref{eq::MSE_L_B}) that are to be used:
%
%For an unblocked experiment:
%\begin{itemize}
%\item Determinant-based $MSE(D)$ criterion:
%\begin{align}
%\label{eq::MSE_D_ms}
%\mbox{minimise }&\left[\left|\bm{X}'_{1}\bm{X}_{1}\right|^{-1/p}F_{p,d;1-\alpha_{DP}}\right]^{\kappa_{DP}} \times \notag \\ &\left[\left|\bm{L}+\frac{\bm{I}_{q}}{\tau^{2}}\right|^{-1/q}F_{q,d;1-\alpha_{LoF}}\right]^{\kappa_{LoF}}\times \notag\\ & \left[|\bm{X}'_{1}\bm{X}_{1}|^{-1}\exp\left(\frac{1}{N}\sum_{i=1}^{N}\log(1+\bm{\tilde{\beta}}_{2i}'\bm{X}_2^{'}\bm{X}_1\bm{M}^{-1}\bm{X}_1^{'}\bm{X}_2\bm{\tilde{\beta}}_{2i})\right)\right]_{.}^{\kappa_{MSE}/p}
%\end{align}
%\item Trace-based $MSE(L)$ criterion:
%\begin{align}
%\label{eq::MSE_L_ms}
%\mbox{minimise} &\left[\frac{1}{p}\mbox{trace}(\bm{WX}'_{1}\bm{X}_{1})^{-1}F_{1,d;1-\alpha_{LP}}\right]^{\kappa_{LP}}\times \notag\\& \left[\frac{1}{q}\mbox{trace}\left(\bm{L}+\frac{\bm{I}_{q}}{\tau^{2}}\right)^{-1}F_{1,d;1-\alpha_{LoF}}\right]^{\kappa_{LoF}}\times 
%\notag\\& \left[\frac{1}{p}\mbox{trace}\{(\bm{X}'_{1}\bm{X}_{1})^{-1}+\tau^2\bm{A}\bm{A}'\}\right]^{\kappa_{MSE}}_{.}
%\end{align}
%\end{itemize} 
% 
%For a blocked experiment:
%\begin{itemize}
%\item Determinant-based $MSE(D)$ criterion:
%\begin{align}
%\label{eq::MSE_D_B_ms}
%\mbox{minimise }&\left[\left|(\bm{X}'_{1}\bm{Q}\bm{X}_{1})^{-1}\right|^{1/p}F_{p,d_B;1-\alpha_{DP}}\right]^{\kappa_{DP}} \times \notag \\ &\left[\left|\bm{\tilde{L}}+\frac{\bm{I}_{q}}{\tau^{2}}\right|^{-1/q}F_{q,d_B;1-\alpha_{LoF}}\right]^{\kappa_{LoF}}\times \notag\\ & \left[|\bm{X}'_{1}\bm{QX}_{1}|^{-1}\exp\left(\frac{1}{N}\sum_{i=1}^{N}\log(1+\bm{\tilde{\beta}}_{2i}'\bm{X}_2^{'}\bm{QX}_1\bm{\tilde{M}}^{-1}\bm{X}_1^{'}\bm{QX}_2\bm{\tilde{\beta}}_{2i})\right)\right]_{.}^{\kappa_{MSE}/p}
%\end{align}
%\item Trace-based $MSE(L)$ criterion:
%\begin{align}
%\label{eq::MSE_L_B_ms}
%\mbox{minimise }&\left[\frac{1}{p}\mbox{trace}(\bm{WX}'_{1}\bm{Q}\bm{X}_{1})^{-1}F_{1,d_B;1-\alpha_{LP}}\right]^{\kappa_{LP}}\times \notag \\&\left[\frac{1}{q}\mbox{trace}\left(\bm{\tilde{L}}+\frac{\bm{I}_{q}}{\tau^{2}}\right)^{-1}F_{1,d_B;1-\alpha_{LoF}}\right]^{\kappa_{LoF}}\times \notag \\& \left[\frac{1}{p}\mbox{trace}\{(\bm{X}'_{1}\bm{QX}_{1})^{-1}+\tau^2[\bm{\tilde{A}}\bm{\tilde{A}}']_{22}\}\right]^{\kappa_{MSE}}_{.}
%\end{align}
%\end{itemize}

>>>>

% Appendix B.  The detailed design search for a multistratum experiment.
